---
title: "Covid"
author: "David"
date: "2025-03-02"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r library}
library(tidyverse)
library(lubridate)
```
## Covid-19 Data
```{r get-data}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_recovered_global.csv")
urls <- str_c(url_in, file_names)
```
```{r import-data}
global_cases <- read.csv(urls[2])
global_deaths <- read.csv(urls[4])
us_cases <- read.csv(urls[1])
us_deaths <- read.csv(urls[3])
global_recovered <- read.csv(urls[5])

```

## Cleaning Data
```{r tidying-global-data}
#dropping Lat,Long columns, pivoting so dates are rows not columns
global_cases <- global_cases %>%
  pivot_longer(cols=
                 -c('Province.State', Lat, Long, 'Country.Region'),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))
#all dates start with the chr X need to remove that
global_cases$date <- global_cases$date %>%
  str_remove("X")

#same for global deaths
global_deaths <- global_deaths %>%
  pivot_longer(cols=
                 -c('Province.State', Lat, Long, 'Country.Region'),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))

global_deaths$date <- global_deaths$date %>%
  str_remove("X")

#global recovered
global_recovered <- global_recovered %>%
  pivot_longer(cols=
                 -c('Province.State', Lat, Long, 'Country.Region'),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_recovered$date <- global_recovered$date %>%
  str_remove("X")
#joining global deaths and cases
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country.Region',
         Province_State = 'Province.State') %>%
  mutate(date = mdy(date))
global <- global %>% filter(cases>0)
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read.csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID,FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```
```{r tidy-us-data}
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to ="cases")
us_cases$date <- us_cases$date %>%
  str_remove("X")

us_cases <- us_cases %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

#us_deaths
us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to ="deaths")
us_deaths$date <- us_deaths$date %>%
  str_remove("X")

us_deaths <- us_deaths %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))
#join
US <- us_cases %>%
  full_join(us_deaths)
US<- US %>% filter(cases>0)

```
```{r visualizing-data}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths=sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000/Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000/Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

global_by_province <- global %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths=sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mil = deaths *1000000/Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mil, Population) %>%
  ungroup()

global_totals <- global_by_province %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mil = deaths * 1000000/Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mil, Population) %>%
  ungroup()

#graph for us
US_totals_graph <- US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x= date, y = cases)) +
  geom_line(aes(color = 'cases')) +
  geom_point(aes(color = 'cases')) +
  geom_line(aes(y=deaths, color = 'deaths')) +
  geom_point(aes(y = deaths, color = 'deaths')) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in US", y = NULL)
print(US_totals_graph)

US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases), 
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases), 
         new_deaths = deaths - lag(deaths))

US_new_totals_graph <- US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x= date, y = new_cases)) +
  geom_line(aes(color = 'new_cases')) +
  geom_point(aes(color = 'new_cases')) +
  geom_line(aes(y=new_deaths, color = 'new_deaths')) +
  geom_point(aes(y = new_deaths, color = 'new_deaths')) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in US", y = 'Amount', x = 'Year')
print(US_new_totals_graph)

US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 *cases/population,
            deaths_per_thou = 1000 *deaths/population) %>%
  filter(cases > 0, population > 0)

#looking at india
india_graph <- global_totals %>%
  filter(cases > 0, Country_Region == c('India')) %>%
  ggplot(aes(x= date, y = cases)) +
  geom_line(aes(color = 'cases')) +
  geom_point(aes(color = 'cases')) +
  geom_line(aes(y=deaths, color = 'deaths')) +
  geom_point(aes(y = deaths, color = 'deaths')) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID19 in India", y = "Amount", x = 'Year')
print(india_graph)

#don't have good province data for india so will just look at all countries
#instead of looking at india by province
country_totals <- global_totals %>%
  filter(Country_Region != 'US') %>%
  group_by(Country_Region)%>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 *cases/population,
            deaths_per_thou = 1000 *deaths/population) %>%
  filter(cases > 0, population > 0)

```
```{r models}
mod <- lm(deaths_per_thou ~ cases_per_thou, data= US_state_totals)
summary(mod)

x_grid <- seq(1,151)
new_df <- tibble(cases_per_thou=x_grid)
US_tot_w_pred <-US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred %>% ggplot() + geom_point(aes(x=cases_per_thou, y = deaths_per_thou),
                                        color = "blue") + geom_point(aes(x=cases_per_thou, y = pred), color = "red")

mod2 <- lm(deaths_per_thou ~ cases_per_thou, data= country_totals)
summary(mod)

x_grid <- seq(1,151)
new_df <- tibble(cases_per_thou=x_grid)
Country_tot_w_pred <- country_totals %>% mutate(pred = predict(mod2))
Country_tot_w_pred %>% ggplot() + geom_point(aes(x=cases_per_thou, y = deaths_per_thou),
                                        color = "blue") + geom_point(aes(x=cases_per_thou, y = pred), color = "red")

```

## Conclusion
We can see that in India and the United States the largest growth in cases
happened in 2020 through 2021. This is to be expected as that is 
when the virus first started to spread globally. We can see that there is a clear
positive linear trend both in the US and globally between cases per thousand and
deaths per thousand. Which is to be expected.

Possible sources of bias include inaccurate data reporting from the government agencies of each country. The COVID-19 pandemic was highly politicized and many
governments did not accurately or efficiently collect and report data on the pandemic.
This would lead to inaccuracies across the dataset which could skew data for certain countries.